Individual and Differential Harm in Redistricting

### 2022-08-30 paper/harm.Rmd
typos

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank Steve Ansolabehere, Kosuke Imai, Gary King, Shiro Kuriwaki, 
      Tyler Simko, Nick Stephanopolous, and 
      participants of the Fall 2021 Harvard American Politics Research Workshop 
      and Imai Research Group for helpful comments and discussion.
      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: apsr
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.

@@

The core idea motivating individual and differential harm is a simple one.
Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
The state's total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two medium-sized cities and a substantial rural population which an overall partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.

@@

Our definition of harm is limited in two regards.
First, it considers only individuals who vote: people who abstain from voting have experience zero harm.
People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
Naturally, implementing this extension in practice would require auxiliary data, such as a large-scale survey, since candidate preferences for non-voters cannot be directly inferred from election returns.

@@

A primary advantage of our proposed harm measures is that they our measures are defined at the individual level,
allowing for the study of redistricting for any defined group memberships.
For political science, two of the most important group memberships are perhaps race and party.
Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.

@@

The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
Plugging in past election results without using an election model places assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.

@@

These values can be better contextualized in comparison the distribution across the sampled plans.
The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.

@@

To apply our proposed framework to study the council district boundaries, must be able to we predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one choice out of many model which could provide estimated vote choices for individuals.
Local survey data with ideological preferences and racial information would be a sufficient alternative.

@@

Modeling future elections poses another set of challenges.
Simple models are easy to build from commonly-available data and are computationally tractable, but are gross simplifications of reality.
More realistic models require more analyst thought and can add significant burden to the calculation of our harm measures.
And as discussed in Section \ref{subsec:harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
Each of these limitations present opportunities for future study.

@@

For each election, we compute the two-party Democratic vote share.
We fit a linear mixed-effects model to these vote shares, with an overall intercept fixed effect, and random effects for districts and election years.
To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
For example, the Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
We do not account for any mid-decade redistricting which may occur as a result of litigation.

@@

The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
These estimates were derived through multilevel regression and post-stratification from survey data, calibrated to actual election results.

@@

Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
To identify the rotation, we place more informative priors on the loadings of certain candidates.
This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
We chose to make align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.

>>>>

---
title: Individual and Differential Harm in Redistricting
authors:
  - name: Cory McCartan
    thanks: |
      To whom correspondence should be addressed.
      The authors thank Steve Ansolabehere, Kosuke Imai, Gary King, Shiro Kuriwaki, 
      Tyler Simko, Nick Stephanopolous, and 
      participants of the Fall 2021 Harvard American Politics Research Workshop 
      and Imai Research Group for helpful comments and discussion.
      Replication code and data are available at <https://github.com/CoryMcCartan/harm-redistricting>.
    department: Department of Statistics
    affiliation: Harvard University
    location: Cambridge, MA
    email: cmccartan@fas.harvard.edu
  - name: Christopher T. Kenny
    department: Department of Government
    affiliation: Harvard University
    location: Cambridge, MA
    email: christopherkenny@fas.harvard.edu
abstract: |
  Social scientists have developed dozens of measures for assessing partisan bias in redistricting.
  But these measures are not easily adapted to other groups, 
  including groups defined by race, class, or geography.
  Nor are they applicable to single- or no-party contexts, such as local redistricting.
  To overcome these limitations, we propose a unified framework of harm for evaluating 
  the impacts of a districting plan on individual voters and the groups to which they belong. 
  We consider a voter harmed if their chosen candidate is not elected under the current plan, 
  but would be under a different plan.
  Harm improves on existing measures by both focusing on the choices of individual voters 
  and directly incorporating counterfactual plans.
  We discuss strategies for estimating harm, and demonstrate the utility of our framework, 
  through analyses of partisan gerrymandering in New Jersey, voting rights 
  litigation in Alabama, and racial dynamics of Boston City Council elections.
keywords:
  - representation
  - redistricting
  - voting rights
  - individual harm
bibliography: references.bib
biblio-style: apsr
output:
  rticles::arxiv_article:
    keep_tex: true
    includes:
      in_header: "header.tex"
    latex_engine: pdflatex
editor_options:
  markdown:
    wrap: sentence
---

@@

The harm framework has important similarities with the common-law approach to voting rights litigation.
To proceed with a legal challenge to an alleged gerrymander, a voter must demonstrate that their individual right to vote has been harmed by the enacted district boundaries [@1964reynolds; @1962baker; @issakarl98]. This requires proof that *but for* the alleged gerrymander, the voter would have been more likely to elect their chosen candidate [@2004vieth; @2018league; @2019lewis].
Indeed, a primary factor in the Supreme Court's ruling that partisan gerrymandering is not justiciable was that existing redistricting measures are not clearly tied to individual injury [@2017gill].
A more detailed discussion of the connections between the proposed framework and existing practices for analyzing racial and partisan gerrymandering may be found in Section \ref{sec:connections}.

@@

The core idea motivating individual and differential harm is a simple one.
Consider *Minissouri*, a stylized representation of Missouri which is shown in Figure \ref{fig:minissouri}.
The state's total population is 180, and it is divided into three districts with population between 54 and 66.
Minissouri has two medium-sized cities and a substantial rural population with an overall partisan breakdown of 55.6% R to 44.4% D.
Voters' preferences are assumed to be stable across elections, and turnout is 100%.
In addition to partisan affiliation indicated by red or blue coloring, each voter also belongs to a racial group indicated by squares and circles.

@@

Our definition of harm is limited in two regards.
First, it considers only individuals who vote: people who abstain from voting have experienced zero harm.
People who do not vote express no formal preference in electoral outcomes, even if they have informal preferences.
The natural theoretical extension of this definition is to then separate individuals' candidate preferences from their vote choice; the latter defines election winners, but the former provides a more-broad harm measure.
Naturally, implementing this extension in practice would require auxiliary data, such as a large-scale survey, since candidate preferences for non-voters cannot be directly inferred from election returns.

@@

A primary advantage of our proposed harm measures is that they are defined at the individual level, allowing for the study of redistricting for any defined group memberships.
For political science, two of the most important group memberships are perhaps race and party.
Even in these well-studied contexts, our measures provide an improvement over existing analytical approaches.

@@

The latter three metrics are grounded in intuition over how to reliably identify a partisan gerrymander, and not in the larger theoretical framework of partisan symmetry [@katz2020].
As such, they make no distinction between the quantity they are trying to estimate and a particular estimator or formula used in calculations.
This is in contrast to partisan symmetry as well as the proposed harm measures, which require practitioners to take the additional step of linking observed election data to the theoretical construct by way of an election model, as discussed in Section \ref{sec:estimating}.[^4]
This is an important conceptual distinction that is often lost in day-to-day applications of various gerrymandering measures.
Plugging in past election results without using an election model assumes too much about the stability of electoral dynamics and consequently can significantly understate the uncertainty in how a particular districting plan will perform into the future.

@@

These values can be better contextualized in comparison to the distribution across the sampled plans.
The Democratic proposal is expected to produce an average of 8.55 Democratic seats under our election model, more than 79% of all sample plans.
The Republican proposal is an outlier, yielding fewer seats on average than every single sampled plan.
This pattern is repeated for all of the four common gerrymandering measures, where the Republican plan is a clear outlier (all $p=0.0002$, except $p=0.002$ for partisan bias) while the Democratic plan is not ($p=0.34$ for the efficiency gap, $0.50$ for declination, $0.88$ for partisan bias, and $0.17$ for the mean-median score).
However, both the partisan bias estimate and the mean-median score again misidentify the direction of the bias of the Republican plan, classifying it as an extreme Democratic outlier.

@@

To apply our proposed framework to study the council district boundaries, we must be able to predict individual voting preferences under counterfactual districting plans.
This is more difficult than in both New Jersey and Alabama, for two reasons.
First, the city council is officially nonpartisan, and city politics are dominated by the Democratic Party, so voting choices are more complex than a simple partisan preference.
To estimate election outcomes, we must map observed support for particular candidates in one election onto counterfactual support for a hypothetical slate in another election.
Second, as in Alabama, we aim to examine the racial impact of the district system, and so must estimate these voting preferences by race.
Even if we could, for instance, divide candidates into a more "liberal" and a more "conservative" group, the vote counts for each group by precinct will not alone show how different racial groups voted.

@@

Like all ecological inference models, ours relies on several unverifiable assumptions.
The latent factor component imposes additional assumptions about the structure of voting in Boston elections.
As Appendix \ref{app:lfei} discusses, these assumptions are almost certainly not *strictly* met.
However, the model is a useful simplification of reality that captures the major electoral dynamics relevant to our question of interest.
For example, using the estimated latent positions and support levels for Michelle Wu and Annissa Essaibi-George, the two mayoral candidates who advanced to the general election, we predict the precinct-level general election turnout and vote shares with considerable accuracy (Spearman correlation $\approx 0.85$, see Figure \ref{fig:boston-gen-pred} in Appendix \ref{app:lfei}).
The accuracy of these validation predictions tracks with previous work using factor models to predict runoff outcomes after a first-stage election [@kamakura2006modeling].
We stress that the latent factor ecological inference model is not necessary for this kind of analysis; it is one out of several approaches which could provide estimated vote choices for individuals.
Local survey data with ideological preferences and racial information would be a sufficient alternative.

@@

Modeling future elections poses another set of challenges.
Simple models are easy to build from commonly-available data and are computationally tractable, but are gross simplifications of reality.
More realistic models require more analyst thought and can add a significant burden to the calculation of our harm measures.
And as discussed in Section \ref{subsec:harm}, multi-candidate elections where voters have strong ordinal preferences provide further challenges that the proposed framework does not fully address.
Each of these limitations present opportunities for future study.

@@

For each election, we compute the two-party Democratic vote share.
We fit a linear mixed-effects model to these vote shares, with an overall intercept fixed effect, and random effects for districts and election years.
To account for the fact that district boundaries change every decade, the district identifiers are built as a concatenation of the state, district number, and redistricting.
For example, Washington's 7th congressional district receives one random effect from 2002-2010 and a different random effect from 2012-2020, since new district boundaries were enacted in the 2010-2012 period.
We do not account for any mid-decade redistricting which may occur as a result of litigation.

@@

The mean support $\mu^{(s)}$ receives an informative prior based on the racially polarized voting estimates of @kuriwaki2021geography, with $\mu^{(s)}\sim\Norm(\logit((0.18, 0.90, 0.35)), (0.03^2, 0.03^2, 0.2^2))$.
These estimates were derived through multilevel regression and poststratification from survey data, calibrated to actual election results.

@@

Like all probabilistic factor models, ours is unidentifiable due to the rotation, location, and scale invariance of the factors in the likelihood.
To identify the location and scale, we place $\Norm(0,1)$ priors on most components of $\vb L$.
To identify the rotation, we place more informative priors on the loadings of certain candidates.
This was accomplished through somewhat of an interactive process, where a preliminary principal components decomposition of the precinct vote shares helped suggest the latent factors that might be found by the model.
We chose to align the first dimension with the Sanders-Biden axis, and chose to make the second dimension roughly orthogonal along the Janey-Buttigieg axis.
This translated to a prior of $\Norm((-1, 0), (0.1^2, 0.25^2))$ for Sanders, $\Norm((1, 0), (1.0^2, 0.5^2))$ for Biden, and $\Norm((-0.25, -1), (0.5^2, 2^2))$ for Janey.
We additionally set the prior correlation between Sanders' and Biden's loadings to be $(-0.5, 0.8)$.
